---
title: "Project"
author: "Gigi Mordock"
format: pdf
editor: source
---

## Section 1: Data Description

```{r}
#| label: load packages
library(baseballr)
```

```{r}
#| label: import data
library(dplyr)
library(Lahman)
teams <- Teams
```

An observation in this data set corresponds to the data from a specific team in a specific year in baseball. There are 3614 observations in this data set with 48 variables. However, I am wanting to reduce this information to the 30 MLB teams over the span of time from 2021-2024 in order to see what data predicts how many wins a team get in a season based on offensive and defensive statistics.  

The Society for American Baseball Research member Sean Lahman collected this data and it is updated after each season. Data is collected up to the 2024 season as recorded in the 2025 version of the database. The data was collected in order to allow people to conduct research projects and simulate games. 

####*DELETE THIS*######
mix of offensive + defensive stats like:
Hitting (offense)
Runs scored (R)
Home runs (HR)
On-base percentage (OBP)
Slugging (SLG)
Batting average (AVG)

Pitching (defense)
Runs allowed (RA)
ERA
Strikeouts (SO)
WHIP

Optional extras
Payroll
Run differential (R - RA)
Walks, hits allowed, etc.

EDA: distributions + scatterplots
Clustering: group teams by style
Prediction: model wins

## Section 2: Data Wrangling and Exploratory Data Analysis


## Section 3: Unsupervised Learning


## Section 4: Supervised Learning

