---
title: "Project"
author: "Gigi Mordock"
format: pdf
editor: source
---

## Section 1: Data Description

```{r}
#| label: load packages
library(baseballr)
library(rsample)
library(dplyr)
library(Lahman)
```

```{r}
#| label: import data
teams <- Teams
teams <- teams |>
  filter(
    yearID >= 2021
  )
teams <- teams |> 
  select(
  yearID, # year played
  name, # name of team
  W, # wins
  G, # games
  R, # runs
  AB, # at bats
  H, # hits
  X2B, # doubles
  X3B, # triples
  HR, # home runs
  BB, # walk
  SO, # strikeout
  SB, # stolen base
  CS, # caught stealing
  HBP, # hit by pitch
  SF, # sacrifice fly
  RA, # opponents runs scored
  ER, # earned runs 
  ERA, # earned run average
  SHO, # shutouts
  SV, # saves 
  IPouts, # outs pitched 
  HA, # hits allowed
  HRA, # homeruns allowed
  BBA, # walks allowed
  E, # errors
  DP # double plays
  )

# figure out WinPct: W/G
# figure out run differential:  R-RA
# figure out WHIP: (BBA+HA)/(Innings Pitched/3)

```

An observation in this data set corresponds to the data from a specific team in a specific year in baseball. There are 3614 observations in the full data set with 48 variables. However, I am wanting to reduce this information to the 30 MLB teams over the span of time from 2021-2025 in order to see what data predicts how many wins a team get in a season based on offensive and defensive statistics. After the changes to adjust for my time period and the necessary variables, there are 150 observations and ******27*CHECK THIS BEFORE TURNING IN**** variables that have to do with offensive and defensive statistics. 

The Society for American Baseball Research member Sean Lahman collected this data and it is updated after each season. Data is collected up to the 2025 season as recorded in the 2026 version of the database. The data was collected in order to allow people to conduct research projects and simulate games. 


```{r}
#| label: split into train and test data
set.seed(2026)
teams_split <- initial_split(teams, prop = .8)
teams_train <- training(teams_split)
teams_test <- testing(teams_split)
```

The data set was split into 80% and 20% of the data, where 80% went into the training set and 20% went into the test set. The training set has 120 observations and the testing set has 30 observations. 


######DELETE EDA: distributions + scatterplots
######DELETE Clustering: group teams by off/def style
######DELETE: Prediction: model wins


## Section 2: Data Wrangling and Exploratory Data Analysis


## Section 3: Unsupervised Learning


## Section 4: Supervised Learning

