---
title: "Project"
author: "Gigi Mordock"
format: pdf
editor: source
---

## Section 1: Data Description

```{r}
#| label: load packages
library(baseballr)
library(rsample)
library(dplyr)
library(Lahman)
library(ggplot2)
library(tidyverse)
library(tidymodels) 
library(kknn)
```

```{r}
#| label: import data
teams <- Teams
teams <- teams |>
  filter(
    yearID >= 2021
  )
teams <- teams |> 
  select(
  yearID, # year played
  name, # name of team
  W, # wins 
  G, # games - 161-162 depending on the season 
  R, # runs 
  AB, # at bats
  H, # hits
  X2B, # doubles
  X3B, # triples
  HR, # home runs
  BB, # walk
  SO, # strikeout
  SB, # stolen base
  CS, # caught stealing
  HBP, # hit by pitch
  SF, # sacrifice fly
  RA, # opponents runs scored
  ER, # earned runs 
  ERA, # earned run average
  SHO, # shutouts
  SV, # saves 
  IPouts, # outs pitched 
  HA, # hits allowed
  HRA, # homeruns allowed
  BBA, # walks allowed
  E, # errors
  DP # double plays
  )
teams <- teams |>
  mutate(
    WinPct = round(W/G, 4), # win percentage
    run_diff = R-RA, # run differential 
    WHIP = round((BBA + HA)/(IPouts/3), 2), # WHIP -> walks, hits per inning pitched
    AVG = round(H/AB, 3), # batting average
    OBP = round((H+BB+HBP)/(AB+BB+HBP+SF), 3), # on-base percentage 
    slugpct = round(((H-X2B-X3B-HR) + 2*X2B + 3*X3B + 4*HR) / AB, 3), # slugging percentage
    OPS = round(OBP + slugpct, 4) # OPS
  )
```

An observation in this data set corresponds to the data from a specific team in a specific year in baseball. There are 3614 observations in the full data set with 48 variables. I restrict the data to the 30 MLB teams over the span of time from 2021-2025 in order to see what data predicts how many wins a team gets in a season based on offensive and defensive statistics. After the changes to adjust for my time period and the necessary variables, there are 150 observations and 27 variables that have to do with offensive and defensive statistics. I have also added 7 more variables that were not included but are calculated from combinations of the given variables. These values were calculated and a small random sample was checked verifying their accuracy with manual checks on random samples.  

The Society for American Baseball Research member Sean Lahman collected this data and it is updated after each season. Data is collected up to the 2025 season as recorded in the 2026 version of the database. The data was collected in order to allow people to conduct research projects and simulate games. 


```{r}
#| label: split into train and test data
set.seed(2026)
teams_split <- initial_split(teams, prop = .8)
teams_train <- training(teams_split)
teams_test <- testing(teams_split)
```

The data set was split into 80% and 20% of the data, where 80% went into the training set and 20% went into the test set. The training set has 120 observations and the testing set has 30 observations. 


######DELETE Clustering: group teams by off/def style
######DELETE: Prediction: model wins

## Section 2: Data Wrangling and Exploratory Data Analysis
Please grade on the additional exploratory Analysis rubric as this data was already very clean. 

I decided to look into some big indicators of success within a baseball team: batting average and earned run average. Batting average is often a great indicator of success from offense and earned run averages are an indicator of defense. This will help me to categorize the groupings of offensive versus defensive teams.   

```{r}
#| label: Boxplot of ERA (2021+)

ggplot(teams, aes(x = ERA)) + 
  geom_histogram(binwidth = 0.20) +
  labs(
    title = "Histogram of Earned Run Average", 
    x = "Earned Run Average"
  )
summary(teams$ERA)
```
The average earned run average for the 5 years and 30 teams was around 4.16, meaning the pitchers allowed about 4.16 runs per 9 innings. This is a great indicator of how batting measures up against pitching. Higher ERA's indicate weaker pitching performance. Teams with a higher ERA might be categorized as having better batting, or just are bad in general (ex. Colorado Rockies in 2025). 


```{r}
#| label: Histogram of Team Batting Average (2021+)

ggplot(
  teams,
  aes(x = AVG)) + 
  geom_histogram(binwidth = 0.005) + 
  labs(
    title = "Distribution of Team Batting Average (2021+)",
    x = "Batting Average",
    y = "Count"
  )

summary(teams$AVG)
```
The mean batting average of the 30 teams over the 5 years was 0.245, meaning that in all, MLB players and teams got a hit 24.5% of the time when they went up to the plate. In context, teams want to aim to be at least at this average in order to be successful. Individual players that are better than this average are considered to be better. The highest average was 0.2760

Continuing on with my exploration of the data, some great comparisons for team performance would be both of the previous variables with the winning percentage to compare how each contributed to the success of the teams. 

```{r}
#| label: Batting Average vs Winning Percentage

ggplot(
  teams, 
  aes(x = AVG, y = WinPct)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(
    title = "Team Batting Average vs Winning Percentage (2021+)",
    x = "Batting Average",
    y = "Winning Percentage"
  )
```
The thing that stands out in this graph is that there is a positive trend with these variables. This pattern shows that teams with better batting averages tend to win a higher proportion of games. 

```{r}
#| label: Earned Run Average vs Winning Percentage
ggplot(
  teams, 
  aes(x = ERA, y = WinPct)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(
    title = "Team ERA vs Winning Percentage (2021+)",
    x = "Earned Run Average",
    y = "Winning Percentage"
  )
```
This relationship has a negative trend meaning that lower ERA (better pitching) is associated with a higher proportion of games won. 

I also wanted to see if I could compare the two variables as well as use the winning percentage to help me distinguish the balance between the two for the teams. 

```{r}
#| label: Batting Average vs. ERA by Winning Percentage

ggplot(
  teams, 
  aes(x = AVG, y = ERA, color = WinPct)) +
  geom_point() +
  labs(
    title = "Batting Average and ERA by Team Success",
    x = "Batting Average",
    y = "Earned Run Average",
    color = "WinPct"
  )
```

With this graph, there is more complexity to the data. The darker the color, the less games won. These darker colors seem to be correlating with the higher ERA and a lower batting average, while the lighter colors seem to correlate with a lower ERA and a higher batting average. This is about what I expected with some odd teams having odd seasons where they lost a lot of games. 

As my follow-up question, I will investigate whether balanced teams have more success compared to just offensive or just defensive teams.

To clasify teams, I define the better versus worse teams for each variable. The above average batting average will indicate a strong offense and a below average ERA will indicate strong defense. 

```{r}
#| label: create team categories

avg_mean <- mean(teams$AVG, na.rm = TRUE)
era_mean <- mean(teams$ERA, na.rm = TRUE)

teams <- teams |>
  mutate(
    team_type = case_when(
      AVG >= avg_mean & ERA <= era_mean ~ "Balanced",
      AVG >= avg_mean & ERA > era_mean ~ "Offense",
      AVG < avg_mean & ERA <= era_mean ~ "Defense",
      TRUE ~ "Weak"
    )
  )
teams|>
  count(team_type) 

teams_train <- teams_train |>
  left_join(
    teams |> select(yearID, name, team_type),
    by = c("yearID", "name")
  )
teams_test <- teams_test |>
  left_join(
    teams |> select(yearID, name, team_type),
    by = c("yearID", "name")
  )

teams|>
  group_by(team_type) |>
  summarise(
    avg_WinPct = mean(WinPct, na.rm = TRUE),
    teams_count = n()
  ) |>
  arrange(desc(avg_WinPct))
```
47 of the teams are balanced, 40 are defense-heavy, 26 are offense-heavy, and 37 are weak in both. With a baseball intuition, we would expect balanced teams to perform the best, which is what our tibble is indeed showing us. Therefore, this supports the idea that teams should be more balanced rather than focusing on just one way of playing. 

## Section 3: Unsupervised Learning

In the context of the data that I already have investigated, k-means clustering seems to be the best fit. I selected performance related statistics because they represent offensive and defensive strengths which aligns with my groupings by playing style. 
```{r}
#| label: choose variables to cluster
cluster_data <- teams_train |>
  select(AVG, ERA, HR, WHIP) |>
  na.omit()

cluster_scaled <- scale(cluster_data)
```

```{r}
#| label: choose the number of clusters

set.seed(2026)
ss <- sapply(1:8, function(k){
  kmeans(cluster_scaled, centers = k, nstart = 25)$tot.withinss
})

plot(1:8, ss, type = "b",
     xlab = "Number of Clusters (k)",
     ylab = "Within-Cluster Sum of Squares",
     main = "Elbow Method for Choosing k Clusters")
```

I want to choose k=3 because the elbow plot shows a clear flattening after 3 clusters meaning that the model would not be as effective with higher numbers of clusters. These groupings may roughly represent stronger and weaker teams, while also separating them by offensive and defensive strength.  

```{r}
#| label: run the k-means clustering

set.seed(2026)
kmeans_teams <- kmeans(cluster_scaled, centers = 3, nstart = 25)
teams_train$cluster <- as.factor(kmeans_teams$cluster)
```

```{r}
#| label: visualize the clusters

ggplot(
  teams_train,
  aes(x = AVG, y = ERA, color = cluster)) + 
  geom_point() +
  labs(
    title = "K-means Clustering of MLB Teams (2021+) (k=3)",
    x = "Batting Average",
    y = "ERA"
  )
```

K-means clustering is a method that groups teams based on how similar their offensive and defensive statistics are. The model compares the teams based on the selected variables. Teams with similar values are going to be grouped together into the same cluster. The goal is to create clusters where teams are as similar as possible within the cluster and incredibly different than the teams in other clusters. 

In order to do this, we have to pick how many groups we want, represented by k = 3. The algorithm randomly places 3 center points called centroids and assigns teams to the closest one based on similarities. The locations are then updated to reflect the average of the group. The process repeats many times until assignments stabilize and stop switching clusters. This helps to discover patterns that may not be obvious when looking at the data initially. 

```{r}
teams_train |>
  group_by(cluster) |>
  summarise(
    avg_WinPct = mean(WinPct),
    avg_AVG = mean(AVG),
    avg_ERA = mean(ERA),
    count = n()
  )
```
Based on the cluster averages, they seem to be separated by playing style. One cluster has higher batting averages and lower ERA's, which correspond with higher winning percentages. Another had weaker overall teams with lower batting averages and higher ERA's. The rest are the teams that were more average. Therefore, the clusters represented low, average, and high performing teams. 
I want to compare this with my previous classification of the Balanced, Offensive, Defensive and Weak teams. 

```{r}
#| label: comparison with 3 variables 

table(teams_train$team_type, teams_train$cluster)

ggplot(
  teams_train,
  aes(x = team_type, fill = cluster)) +
  geom_bar(position = "fill") +
  labs(
    title = "Comparison of Categories vs. K-Means Clustering",
    x = "Theoretical Team Type",
    y = "Proportion",
    fill = "Cluster"
  )
```
This table shows how the categories I created earlier compare with the clusters assigned by the k-means clusters. This allows us to see whether the groups matched the clusters. If many teams appear with the same cluster, this suggests that the original classifications and clusters agree. Overlap across multiple clusters means that there is more to the categorization. 
I thought it was interesting to see that the balanced teams were mostly in clusters 1 and 2 while the weak teams were mostly in 2 and 3. Unsurprisingly, there was none of cluster 3 in the balanced category, and none of cluster 1 in the weak category. If I had chosen 4 clusters, I predict that the groupings would be more balanced between the clusters with more in the weak type. 

## Section 4: Supervised Learning

For my models, I want to predict how many games a team would win using a k-nearest neighbors regression and a multiple linear regression model. KNN predicts the number of wins by finding statistically similar teams in the training data and averaging the total wins of the k closest teams based on the variables selected. We have to normalize these statistics because the k-nearest neighbors method depends on calculating the distance between the teams in terms of statistical distance. As you will see below, the assumption is reasonable because similar statistics tended to correlate to similar win numbers. 
```{r}
#| label: select predictors

#split into train and test new 
set.seed(2026)
model_train <- teams_train |>
  select(W, AVG, ERA, HR, WHIP, run_diff)
model_test <- teams_test |>
  select(W, AVG, ERA, HR, WHIP, run_diff)

knn_model <- nearest_neighbor(
  mode = "regression",
  neighbors = tune()
)

# recipe 
knn_recipe <- recipe(
  W ~ ., 
  data = model_train
) |>
  step_normalize(all_numeric_predictors())

#workflow
knn_wflow <- workflow() |>
  add_model(knn_model) |>
  add_recipe(knn_recipe)

# cross validation
set.seed(2026)
cv_folds <- vfold_cv(model_train, v = 5)

# tuning 
knn_grid <- tibble(
  neighbors = seq(1, 25, by = 2)
)

knn_tune <- tune_grid(
  knn_wflow,
  resamples = cv_folds, 
  grid = knn_grid,
  metrics = metric_set(rmse)
)

# select best model and k
best_k <- select_best(knn_tune, metric = "rmse")
final_knn <- finalize_workflow(
  knn_wflow, 
  best_k
)

# fit the model 
final_knn_fit <- fit(final_knn, data = model_train)

# predictions on test set
knn_preds <- predict(final_knn_fit, model_test) |>
  bind_cols(model_test)

head(knn_preds)

# evaluate performance of the model 
rmse(knn_preds, truth = W, estimate = .pred)

```
The test rmse means that my model's prediction of team wins are off by about 4.5 wins each season. In the context of MLB, 162 games are played and wins typically vary across teams by 10s. Therefore, the model performs reasonably well in capturing overall team strength based on offensive and pitching statistics.  

For this process, we tuned the number of neighbors (k) to control how many similar teams are used to make a prediction for another given team. For example, our k was 7 so the model chose the 7 most similar teams in the training data and averaged the wins to create the prediction of how many wins the team would have. In general, k controls the bias vs variance of the model. If k is small, the model is too flexible and follows the training data, which means low bias and high variance. If k is large, the model is more stable but has higher bias and lower variance since there might be some patterns missed. 

Next, we are going to fit a multiple linear regression model. I used multiple linear regression because I was curious about if standard statistics can directly explain how many wins a team will get in a linear way. Multiple linear regression works by creating a single equation that represents the relationship between the predictors and the response variable (wins). It estimates how much each predictor affects the number of wins while keeping the rest of the predictors the same. This method assumes that the relationship between the predictors and response variable is approximately linear. Another assumption is that the observations are independent and that the predictors are not excessively correlated with each other. These seem to hold with my EDA and considering the variables I chose. 

```{r}
#| label: fit the model

lm_fit <- linear_reg() |>
  set_engine("lm") |>
  fit(
    W ~ AVG + ERA + HR + WHIP + run_diff,
    data = model_train
  )

tidy(lm_fit)
# The negative coefficient for ERA suggests that teams allowing fewer earned runs tend to win more games. 

lm_preds <- predict(lm_fit, model_test) |>
  bind_cols(model_test) 
head(lm_preds)
rmse(lm_preds, truth = W, estimate = .pred)

set.seed(2026)
baseball_folds <- vfold_cv(
  model_train,
  v = 5,
  repeats = 2
)

lm_model <-linear_reg() |>
  set_engine("lm")

lm_wflow <- workflow() |>
  add_model(lm_model) |>
  add_formula(W ~ AVG + ERA + HR + WHIP + run_diff)

lm_cv <- fit_resamples(
  lm_wflow, 
  resamples = baseball_folds,
  metrics = metric_set(rmse, rsq)
)

# extract metrics for KNN & linear regression
collect_metrics(knn_tune)
collect_metrics(lm_cv)
```
Using 5-fold cross-validation, the KNN model achieved an average RMSE of 5.11 wins with a standard deviation of 0.37, indicating a moderate variability among folds. The linear regression model has a cross-validated RMSE of 4.37 wins with a standard deviation of 0.21 and an average R^2 of 0.89 with a standard deviation of 0.01. This means that there is a strong accuracy and low variability across the folds. 
```{r}
#| label: rmse values

knn_rmse <- rmse(knn_preds, truth = W, estimate = .pred)
lm_rmse <- rmse(lm_preds, truth = W, estimate = .pred)
knn_rmse
lm_rmse
```

Comparing both methods, the models worked reasonably well at predicting how many wins a team would get. The multiple linear regression model would be preferred because it produced a lower RMSE of 3.59 wins compared to the KNN model RMSE of 4.59 wins. 

## Conclusion
This project explored how offense and defense statistics influence a team's success in Major League Baseball from 2021 to 2025. Stronger batting averages and lower ERA values were associated with higher winning percentages, as discovered through exploratory data analysis. Using k-means clustering, teams were grouped with similar playing styles and levels of success. Both K-nearest neighbors and multiple linear regression showed that team wins can be predicted reasonably well using team statistics. Overall, the project demonstrated that well balanced teams perform the best, and teams weaker offensively or defensively may benefit from improving those weaker areas in order to become more balanced.  