---
title: "Project"
author: "Gigi Mordock"
format: pdf
editor: source
---

## Section 1: Data Description

```{r}
#| label: load packages
library(baseballr)
library(rsample)
library(dplyr)
library(Lahman)
```

```{r}
#| label: import data
teams <- Teams
teams <- teams |>
  filter(
    yearID >= 2021
  )
teams <- teams |> 
  select(
  yearID, # year played
  name, # name of team
  W, # wins 
  G, # games - 161-162 depending on the season 
  R, # runs 
  AB, # at bats
  H, # hits
  X2B, # doubles
  X3B, # triples
  HR, # home runs
  BB, # walk
  SO, # strikeout
  SB, # stolen base
  CS, # caught stealing
  HBP, # hit by pitch
  SF, # sacrifice fly
  RA, # opponents runs scored
  ER, # earned runs 
  ERA, # earned run average
  SHO, # shutouts
  SV, # saves 
  IPouts, # outs pitched 
  HA, # hits allowed
  HRA, # homeruns allowed
  BBA, # walks allowed
  E, # errors
  DP # double plays
  )
teams <- teams |>
  mutate(
    WinPct = round(W/G, 4), # win percentage
    run_diff = R-RA, # run differential 
    WHIP = round((BBA + HA)/(IPouts/3), 2) # WHIP -> walks, hits per inning pitched
  )

```

An observation in this data set corresponds to the data from a specific team in a specific year in baseball. There are 3614 observations in the full data set with 48 variables. However, I am wanting to reduce this information to the 30 MLB teams over the span of time from 2021-2025 in order to see what data predicts how many wins a team get in a season based on offensive and defensive statistics. After the changes to adjust for my time period and the necessary variables, there are 150 observations and 27 variables that have to do with offensive and defensive statistics. I have also added 3 more variables that were not included but are calculated from combinations of the given variables. 

The Society for American Baseball Research member Sean Lahman collected this data and it is updated after each season. Data is collected up to the 2025 season as recorded in the 2026 version of the database. The data was collected in order to allow people to conduct research projects and simulate games. 


```{r}
#| label: split into train and test data
set.seed(2026)
teams_split <- initial_split(teams, prop = .8)
teams_train <- training(teams_split)
teams_test <- testing(teams_split)
```

The data set was split into 80% and 20% of the data, where 80% went into the training set and 20% went into the test set. The training set has 120 observations and the testing set has 30 observations. 


######DELETE EDA: distributions + scatterplots
######DELETE Clustering: group teams by off/def style
######DELETE: Prediction: model wins

## Section 2: Data Wrangling and Exploratory Data Analysis
Please grade on the additional exploratory Analysis rubric as this data was already very clean. 

#Produce univariate summaries and graphs of at least two variables (features and/or outcomes)
#Produce at least two graphs showing bivariate or multivariate trends (or lack thereof)
#Write a narrative describing what was explored, why, and what was found (in context). Interleave the code and graphs with the narrative in the appropriate places. Include follow-up questions and answers as appropriate to the narrative. 
#Explore in-depth one or more follow-up questions as appropriate to the original narrative.
#Write a narrative describing what you followed up on, why, and what you found (in context). Interleave the code and graphs with the narrative in the appropriate places. 

## Section 3: Unsupervised Learning
#Select one of the following techniques: Principal Component Analysis, k-means clustering, or model-based clustering. You may additionally choose hierarchical clustering if you think the training set is small enough that you can get a reasonable-looking dendrogram. 
#Perform the technique on the training set. Select the number of PCs or number of clusters. Justify your selection.
#Explain the general concepts/procedure behind the algorith you are implementing. Attempt as best you can to write your explanation in language that someone without much statistical knowledge can understand.
#Produce appropriate graphs showing the results of your technique. Explain what your chosen PCs (at least the first two) or chosen clusters (at least two) represent in context. 


## Section 4: Supervised Learning

